---
title: "DGP Dossier: `r params$meta$dgp_id`"
output: html_document
params:
  meta: NULL
  dgp_id: "synth_baseline"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(CausalStress)
library(ggplot2)
have_kableextra <- requireNamespace("kableExtra", quietly = TRUE)
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork_available <- TRUE
} else {
  patchwork_available <- FALSE
}
```

## 1. Identity & Status

```{r meta_table, echo=FALSE, results='asis'}
stars <- params$meta$difficulty$stars %||% 0
active_star   <- '<span style="color: #f1c40f;">★</span>'
inactive_star <- '<span style="color: #e0e0e0;">★</span>'
difficulty_val <- paste0(
  strrep(active_star, max(0, stars)),
  strrep(inactive_star, max(0, 5 - stars)),
  " (", stars, "/5)"
)
prof <- params$meta$stress_profile

get_badge_color <- function(val) {
  val <- tolower(as.character(val))
  if (val %in% c("linear", "gaussian", "strong", "high", "easy", "stable", "att")) return("success")
  if (val %in% c("moderate", "medium", "experimental")) return("warning")
  if (val %in% c("nonlinear", "heavytail", "heavy", "poor", "severe", "deprecated", "heterogeneous")) return("danger")
  "secondary"
}

stress_val <- if (is.null(prof)) {
  "N/A"
} else {
  badge_vec <- mapply(
    function(k, v) {
      cls <- get_badge_color(v)
      sprintf('<span class="badge bg-%s" style="margin-right:4px;">%s: %s</span>', cls, k, v)
    },
    names(prof),
    unlist(prof),
    USE.NAMES = FALSE
  )
  paste0("&nbsp;", paste(badge_vec, collapse = ""))
}

status_raw <- params$meta$status %||% "N/A"
status_class <- get_badge_color(status_raw)
status_val <- sprintf('<span class="badge bg-%s">%s</span>', status_class, status_raw)

version_val <- params$meta$version %||% "N/A"
dgp_id_val <- params$meta$dgp_id %||% "N/A"
if (is.character(dgp_id_val)) {
  dgp_id_val <- dgp_id_val
}

meta_df <- tibble::tibble(
  Attribute = c("DGP ID", "Version", "Status", "Difficulty", "Stress Profile"),
  Value = c(
    dgp_id_val,
    version_val,
    status_val,
    difficulty_val,
    stress_val
  )
)

knitr::kable(meta_df, format = "html", col.names = NULL, escape = FALSE) %>%
  {
    if (have_kableextra) {
      kableExtra::kable_styling(
        .,
        bootstrap_options = c("hover", "condensed"),
        full_width = TRUE,
        position = "left"
      ) %>%
        kableExtra::column_spec(1, bold = TRUE, width = "150px")
    } else {
      .
    }
  }
```

## 2. What This DGP Stresses (Intent)

Baseline sanity check: if an estimator fails here, it is fundamentally broken. Intended to confirm correct implementation under ideal linear, well-behaved conditions.

## 3. Identification Assumptions (Explicit)

-   Selection on observables holds.
-   Overlap is moderate (propensity bounded).
-   SUTVA holds.

## 4. Mathematical Specification

Covariates: $X_1, X_2, X_3, X_4, X_5 \sim \mathcal{N}(0, 1)$ independently. Only $X_1, X_2$ drive outcomes and treatment.

**Outcome (control):** $$\mu_0(X) = 1 + X_1 + 0.5 X_2$$

**Propensity:** $$p(X) = \text{expit}(0.5 X_1 - 0.5 X_2)$$

**Treatment effect:** $$\tau(X) = 1 + 0.5 X_1$$

Treated outcome: $Y_1 = Y_0 + \tau(X)$ with Gaussian noise $\varepsilon \sim \mathcal{N}(0, 0.5)$.

## 5. Oracle Truth Definition

-   True ATT is calculated as the sample mean of `structural_te` among treated units.
-   No Monte Carlo approximation needed for ATT (exact sample truth).

## 6. Visual Diagnostics (n = 5000)

```{r visuals, fig.width=10, fig.height=7}
set.seed(123)
dgp <- dgp_synth_baseline(n = 5000, seed = 123)
df <- dgp$df

p1 <- ggplot(df, aes(x = p, fill = factor(w))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(title = "Propensity Distribution (Moderate Overlap)",
       x = "p(X)", y = "Count", fill = "W") +
  theme_minimal()

p2 <- ggplot(df, aes(x = X1, y = structural_te)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = expression(paste("Treatment Effect: ", tau(X), " = 1 + 0.5·X1")),
       x = "X1", y = "structural_te") +
  theme_minimal()

p3 <- ggplot(df, aes(x = X1, y = X2, color = factor(w))) +
  geom_point(alpha = 0.4) +
  labs(title = "Covariate Balance", x = "X1", y = "X2", color = "W") +
  theme_minimal()

# Plot the True QST curve (stored in dgp_viz$true_qst)
p4 <- ggplot(dgp$true_qst, aes(x = tau, y = value)) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "gray70") + # Mean ATT reference
  geom_line(color = "#2c3e50", size = 1) +
  labs(
    title = "True Quantile Shift (QST)",
    subtitle = "Solid line: True effect at each quantile | Dashed: Mean ATT",
    x = "Quantile (tau)",
    y = "Treatment Effect"
  ) +
  theme_minimal()

if (patchwork_available) {
  library(patchwork)
  (p1 + p2) / (p3 + p4)
} else {
  print(p1)
  print(p2)
  print(p3)
  print(p4)
}
```

## 7. Empirical Validation

We expect correctly specified OLS (`lm_att`) and the oracle to be essentially unbiased on this baseline.

```{r empirical, cache=TRUE}
run_fast <- function(est_id) {
  cs_run_seeds(
    dgp_id        = "synth_baseline",
    estimator_id  = est_id,
    n             = 5000,
    seeds         = 1:10,
    bootstrap     = FALSE,
    parallel      = FALSE,
    show_progress = FALSE
  )
}

res_lm     <- run_fast("lm_att")
res_oracle <- run_fast("oracle_att")

summary_tbl <- dplyr::bind_rows(res_lm, res_oracle) %>%
  cs_tidy() %>%
  dplyr::group_by(estimator_id) %>%
  dplyr::summarise(
    mean_bias = mean(att_error, na.rm = TRUE),
    rmse      = sqrt(mean(att_error^2, na.rm = TRUE)),
    .groups   = "drop"
  )

knitr::kable(summary_tbl, digits = 3, caption = "Empirical validation (n=5000, seeds=1:10)")
```

## 8. Failure Mode Summary

-   Naive estimators (unadjusted) fail due to confounding.
-   L2 estimators succeed (linear surface).
-   Propensity estimators succeed (moderate overlap).

## 9. Implementation Reference

-   Code: `R/dgp-synth-baseline.R` (v1.3.0)
-   Registry: `cs_dgp_registry()` entry for `synth_baseline`
-   Oracle: `true_att` computed from `structural_te` among treated
-   Airlock: oracle columns removed unless explicitly allowed

## 10. Validation Checklist

-   [x] Math matches code
-   [x] Oracle matches definition
-   [x] Visual diagnostics show intended properties
-   [x] Empirical validation (lm_att vs oracle_att) passes
-   [x] Stress profile/difficulty tags set

## 11. Changelog

-   v1.0 dossier: Initial dossier following Standard Template v1.0
