---
title: "DGP Dossier: `r params$meta$dgp_id`"
output: html_document
params:
  meta: NULL
  dgp_id: "synth_placebo_kangschafer"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(CausalStress)
library(ggplot2)
have_kableextra <- requireNamespace("kableExtra", quietly = TRUE)
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork_available <- TRUE
} else {
  patchwork_available <- FALSE
}
```

## 1. Identity & Status

```{r meta_table, echo=FALSE, results='asis'}
stars <- params$meta$difficulty$stars %||% 0
active_star   <- '<span style="color: #f1c40f;">&#9733;</span>'
inactive_star <- '<span style="color: #e0e0e0;">&#9734;</span>'
difficulty_val <- paste0(
  strrep(active_star, max(0, stars)),
  strrep(inactive_star, max(0, 5 - stars)),
  " (", stars, "/5)"
)
prof <- params$meta$stress_profile

get_badge_color <- function(val) {
  val <- tolower(as.character(val))
  if (val %in% c("linear", "gaussian", "strong", "high", "easy", "stable", "att")) return("success")
  if (val %in% c("moderate", "medium", "experimental")) return("warning")
  if (val %in% c("nonlinear", "heavytail", "heavy", "poor", "severe", "deprecated", "heterogeneous")) return("danger")
  "secondary"
}

stress_val <- if (is.null(prof)) {
  "N/A"
} else {
  badge_vec <- mapply(
    function(k, v) {
      cls <- get_badge_color(v)
      sprintf('<span class="badge bg-%s" style="margin-right:4px;">%s: %s</span>', cls, k, v)
    },
    names(prof),
    unlist(prof),
    USE.NAMES = FALSE
  )
  paste0("&nbsp;", paste(badge_vec, collapse = ""))
}

status_raw <- params$meta$status %||% "N/A"
status_class <- get_badge_color(status_raw)
status_val <- sprintf('<span class="badge bg-%s">%s</span>', status_class, status_raw)

version_val <- params$meta$version %||% "N/A"
dgp_id_val <- params$meta$dgp_id %||% "N/A"
if (is.character(dgp_id_val)) {
  dgp_id_val <- dgp_id_val
}

meta_df <- tibble::tibble(
  Attribute = c("DGP ID", "Version", "Status", "Difficulty", "Stress Profile"),
  Value = c(
    dgp_id_val,
    version_val,
    status_val,
    difficulty_val,
    stress_val
  )
)

knitr::kable(meta_df, format = "html", col.names = NULL, escape = FALSE) %>%
  {
    if (have_kableextra) {
      kableExtra::kable_styling(
        .,
        bootstrap_options = c("hover", "condensed"),
        full_width = TRUE,
        position = "left"
      ) %>%
        kableExtra::column_spec(1, bold = TRUE, width = "150px")
    } else {
      .
    }
  }
```

## 2. What This DGP Stresses (Intent)

This is the canonical Kang-Schafer misspecification trap adapted as a sharp-null placebo.
It is designed to create a practical violation of positivity: the true propensity is
generated from latent variables, but analysts only observe nonlinear transforms.
A logistic regression on the observed covariates looks reasonable yet is slightly
misspecified, leading to unstable inverse probability weights.

Reference: *Kang, J. D. Y., & Schafer, J. L. (2007). Demystifying Double Robustness:
A Comparison of Alternative Strategies for Estimating a Population Mean from Incomplete Data.
Statistical Science, 22(4), 523-539.*

Estimators that report a nonzero effect here fail the Gatekeeper check.

## 3. Identification Assumptions (Explicit)

- Selection on observables holds in the latent Z space.
- Observed covariates are nonlinear transforms of Z, creating misspecification.
- Sharp null: Y1 == Y0 pathwise.

## 4. Mathematical Specification

Latent covariates:
\[
Z = (Z_1, Z_2, Z_3, Z_4)^\top \sim \mathcal{N}_4(0, I_4)
\]

Observed covariates (nonlinear transforms):
\[
X_1 = \exp(Z_1/2),\quad X_2 = \frac{Z_2}{1 + \exp(Z_1)} + 10,
\]
\[
X_3 = (Z_1 Z_3 / 25 + 0.6)^3,\quad X_4 = (Z_2 + Z_4 + 20)^2
\]

Propensity in latent space:
\[
p(Z) = \text{expit}(-Z_1 + 0.5 Z_2 - 0.25 Z_3 - 0.1 Z_4)
\]

Outcome in latent space:
\[
\mu_0(Z) = 210 + 27.4 Z_1 + 13.7 Z_2 + 13.7 Z_3 + 13.7 Z_4
\]

Sharp null effect:
\[
\tau(Z) \equiv 0,\quad Y_1 = Y_0
\]

**Noise:**
\[
\varepsilon \sim \mathcal{N}(0, 1^2)
\]

## 5. Oracle Truth Definition

- True ATT is exactly 0 (sharp null).
- True QST is identically 0 on the oracle tau grid.

## 6. Visual Diagnostics (n = 5000)

```{r visuals, fig.width=10, fig.height=7}
set.seed(123)
dgp <- dgp_synth_placebo_kangschafer(n = 5000, seed = 123)
df <- dgp$df

p1 <- ggplot(df, aes(x = p, fill = factor(w))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(title = "Propensity Distribution (Practical Positivity Violation)",
       subtitle = "Histogram shows true p(Z). Analysts observing X face misspecification.",
       x = "p(Z)", y = "Count", fill = "W") +
  theme_minimal()

p2 <- ggplot(df, aes(x = X1, y = structural_te)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Treatment Effect: tau(X) = 0",
       x = "X1", y = "structural_te") +
  theme_minimal()

p3 <- ggplot(df, aes(x = X1, y = X2, color = factor(w))) +
  geom_point(alpha = 0.4) +
  labs(title = "Covariate Overlap (Joint Support)", x = "X1", y = "X2", color = "W") +
  theme_minimal()

p4 <- ggplot(dgp$true_qst, aes(x = tau, y = value)) +
  geom_hline(yintercept = dgp$true_att, linetype = "dashed", color = "gray70") +
  geom_line(color = "#2c3e50", size = 1) +
  labs(
    title = "True Quantile Shift (QST)",
    subtitle = "Flat line at zero (sharp null)",
    x = "Quantile (tau)",
    y = "Treatment Effect"
  ) +
  theme_minimal()

if (patchwork_available) {
  library(patchwork)
  (p1 + p2) / (p3 + p4)
} else {
  print(p1)
  print(p2)
  print(p3)
  print(p4)
}
```

## 7. Empirical Validation

Oracle should be zero. IPW is expected to be noisy or biased under misspecification.

```{r empirical, cache=TRUE}
run_fast <- function(est_id) {
  cs_run_seeds(
    dgp_id        = "synth_placebo_kangschafer",
    estimator_id  = est_id,
    n             = 1000,
    seeds         = 1:20,
    bootstrap     = FALSE,
    parallel      = FALSE,
    show_progress = FALSE
  )
}

res_oracle <- run_fast("oracle_att")
res_ipw <- tryCatch(run_fast("ipw_att"), error = function(e) NULL)

summary_tbl <- dplyr::bind_rows(res_oracle, res_ipw) %>%
  cs_tidy() %>%
  dplyr::group_by(estimator_id) %>%
  dplyr::summarise(
    mean_bias = mean(att_error, na.rm = TRUE),
    rmse      = sqrt(mean(att_error^2, na.rm = TRUE)),
    .groups   = "drop"
  )

knitr::kable(summary_tbl, digits = 3, caption = "Empirical validation (n=1000, seeds=1:20)")
```

**Interpretation:** The oracle remains at zero as expected. IPW can show erratic
behavior because the propensity model is slightly misspecified in the observed
covariate space, producing unstable weights.

## 8. Failure Mode Summary

- IPW and DR methods are fragile to mild misspecification here (weight explosion).
- Any detected nonzero effect indicates failure of the gatekeeper check.

## 9. Implementation Reference

- Generator: `R/dgp-synth-placebo-kangschafer.R`
- Registry: `cs_dgp_registry()` entry for `synth_placebo_kangschafer`
- Oracle truth: `cs_get_oracle_qst("synth_placebo_kangschafer", version = "1.4.0")`

## 10. Validation Checklist

- [x] Sharp null (Y1 == Y0) enforced
- [x] Nonlinear transforms of latent Z documented
- [x] Propensity histogram shows skew/edge mass
- [x] Oracle returns zero effect
- [x] IPW fragility explained

## 11. Changelog

- v1.4.0 (2026-01-06): Revised dossier with Kang-Schafer citation and misspecification narrative.

## Appendix: Implementation

```{r impl_code, echo=FALSE, comment=''}
# Dynamically retrieve the specific versioned function for this dossier
dgp_base <- paste0("dgp_", params$dgp_id)
dgp_ver  <- if (!is.null(params$meta$version)) {
  paste0(dgp_base, "_v", gsub("\\.", "", params$meta$version))
} else {
  dgp_base
}

if (exists(dgp_ver)) {
  cat(sprintf("Source code for: %s\n\n", dgp_ver))
  print(get(dgp_ver))
} else if (exists(dgp_base)) {
  cat(sprintf("Source code for: %s (Wrapper)\n\n", dgp_base))
  print(get(dgp_base))
} else {
  cat("Source code not found in loaded namespace.")
}
```
