---
title: "DGP Dossier: `r params$meta$dgp_id`"
format: html
params:
  meta: NULL
  dgp_id: "synth_heavytail"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(CausalStress)
library(ggplot2)
library(dplyr)
have_kableextra <- requireNamespace("kableExtra", quietly = TRUE)
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork_available <- TRUE
} else {
  patchwork_available <- FALSE
}
```

## 1. Identity & Status

```{r meta_table, echo=FALSE, results='asis'}
stars <- params$meta$difficulty$stars %||% 0
active_star   <- '<span style="color: #f1c40f;">&#9733;</span>'
inactive_star <- '<span style="color: #e0e0e0;">&#9734;</span>'
difficulty_val <- paste0(
  strrep(active_star, max(0, stars)),
  strrep(inactive_star, max(0, 5 - stars)),
  " (", stars, "/5)"
)
prof <- params$meta$stress_profile

get_badge_color <- function(val) {
  val <- tolower(as.character(val))
  if (val %in% c("linear", "gaussian", "strong", "high", "easy", "stable", "att")) return("success")
  if (val %in% c("moderate", "medium", "experimental")) return("warning")
  if (val %in% c("nonlinear", "heavytail", "heavy", "poor", "severe", "deprecated", "heterogeneous")) return("danger")
  "secondary"
}

stress_val <- if (is.null(prof)) {
  "N/A"
} else {
  badge_vec <- mapply(
    function(k, v) {
      cls <- get_badge_color(v)
      sprintf('<span class="badge bg-%s" style="margin-right:4px;">%s: %s</span>', cls, k, v)
    },
    names(prof),
    unlist(prof),
    USE.NAMES = FALSE
  )
  paste0("&nbsp;", paste(badge_vec, collapse = ""))
}

status_raw <- params$meta$status %||% "N/A"
status_class <- get_badge_color(status_raw)
status_val <- sprintf('<span class="badge bg-%s">%s</span>', status_class, status_raw)

version_val <- params$meta$version %||% "N/A"
dgp_id_val <- params$meta$dgp_id %||% "N/A"
if (is.character(dgp_id_val)) {
  dgp_id_val <- dgp_id_val
}

meta_df <- tibble::tibble(
  Attribute = c("DGP ID", "Version", "Status", "Difficulty", "Stress Profile"),
  Value = c(
    dgp_id_val,
    version_val,
    status_val,
    difficulty_val,
    stress_val
  )
)

knitr::kable(meta_df, format = "html", col.names = NULL, escape = FALSE) %>%
  {
    if (have_kableextra) {
      kableExtra::kable_styling(
        .,
        bootstrap_options = c("hover", "condensed"),
        full_width = TRUE,
        position = "left"
      ) %>%
        kableExtra::column_spec(1, bold = TRUE, width = "150px")
    } else {
      .
    }
  }
```

## 2. What This DGP Stresses (Intent)

**Purpose:** Penalize estimators that rely on L2 loss under heavy-tailed noise.

This DGP is identical to the baseline linear design except for the outcome noise,
which is a Gaussian/Cauchy mixture. The Cauchy component makes the population mean
of observed outcomes undefined. Consequently, a conventional superpopulation mean
potential-outcome ATT does not exist, and estimators that rely on L2 loss or outcome
means do not converge to such an estimand.

CausalStress deliberately keeps running ATT estimators because their breakdown is
the evidence this boundary DGP is designed to produce. The package's `true_att`
is a governed finite-sample structural signal anchor, not a conventional
superpopulation mean potential-outcome ATT. It must not be used for an ATT
estimator shootout. QST is the well-defined distributional comparison target in
this regime.

## 3. Identification Assumptions (Explicit)

- Unconfoundedness holds by construction (treatment is assigned by a known
  propensity score).
- Overlap is moderate (`p(X)` bounded away from 0 and 1 for most of the mass).
- The governed structural signal anchor is the finite-sample mean of `tau(X_i)`
  over the run's treated units (Constitution Article I), independent of realized
  noise. Because the population outcome means do not exist, this is not a
  conventional superpopulation mean potential-outcome ATT.

## 4. Mathematical Specification

**Covariates**
\[
X \sim \mathcal{N}(0, I_5)
\]

**Propensity**
\[
p(X) = \text{expit}(0.5 X_1 - 0.5 X_2)
\]

**Baseline Outcome and Treatment Effect**
\[
\mu_0(X) = 1 + X_1 + 0.5 X_2
\]
\[
\tau(X) = 1 + 0.5 X_1
\]

**Noise Mixture (Heavy Tails)**
\[
\varepsilon \sim 0.8 \cdot \mathcal{N}(0, 0.5^2) + 0.2 \cdot \text{Cauchy}(0, 1)
\]
where \(\varepsilon_0\) and \(\varepsilon_1\) are drawn independently.

**Moment note:** The Cauchy component has undefined mean and variance. Therefore the
population mean of realized outcomes is undefined, and any L2/MSE estimator is not
well-posed in the population.

**Potential Outcomes**
\[
Y_0 = \mu_0(X) + \varepsilon_0,\quad
Y_1 = \mu_0(X) + \tau(X) + \varepsilon_1
\]

## 5. Oracle Truth Definition

- **Finite-sample structural signal anchor (`true_att`):**
  \(N_{treated}^{-1}\sum_{i:W_i=1}\tau(X_i)\), computed from the deterministic
  signal \(\tau(X)\). This governed diagnostic anchor is finite even though the
  conventional superpopulation mean potential-outcome ATT is not.
- **QST (Distributional):** computed on the canonical tau grid via oracle Monte Carlo
  as specified in `cs_get_oracle_qst()`.

## 6. Visual Diagnostics (n = 5000)

```{r visuals, fig.width=10, fig.height=7}
set.seed(123)
dgp <- dgp_synth_heavytail(n = 5000, seed = 123)
df <- dgp$df

p1 <- ggplot(df, aes(x = p, fill = factor(w))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(title = "Propensity Distribution (Moderate Overlap)",
       x = "p(X)", y = "Count", fill = "W") +
  theme_minimal()

p2 <- ggplot(df, aes(x = X1, y = structural_te)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = expression(paste("Treatment Effect: ", tau(X), " = 1 + 0.5*X1")),
       x = "X1", y = "structural_te") +
  theme_minimal()

p3 <- ggplot(df, aes(x = X1, y = X2, color = factor(w))) +
  geom_point(alpha = 0.4) +
  labs(title = "Covariate Overlap (Joint Support)", x = "X1", y = "X2", color = "W") +
  theme_minimal()

p4 <- ggplot(dgp$true_qst, aes(x = tau, y = value)) +
  geom_hline(yintercept = dgp$true_att, linetype = "dashed", color = "gray70") +
  geom_line(color = "#2c3e50", size = 1) +
  labs(
    title = "True Quantile Shift (QST)",
    subtitle = "Solid line: True effect at each quantile | Dashed: Mean ATT",
    x = "Quantile (tau)",
    y = "Treatment Effect"
  ) +
  theme_minimal()

if (patchwork_available) {
  library(patchwork)
  (p1 + p2) / (p3 + p4)
} else {
  print(p1)
  print(p2)
  print(p3)
  print(p4)
}
```

## 7. Empirical Validation

We run light validation with `n = 1000` and `seeds = 1:20`. The goal is to show
extreme sensitivity of OLS under L2 loss, not to rank ATT estimators. We report
finite-replicate diagnostics relative to the structural signal anchor, centered
on medians and a quantile; the sample maximum is descriptive only. These are not
conventional ATT bias or RMSE. We do **not** run heavier robust estimators here
to keep doc builds fast.

```{r empirical, cache=FALSE}
run_fast <- function(est_id) {
  cs_run_seeds(
    dgp_id        = "synth_heavytail",
    estimator_id  = est_id,
    n             = 1000,
    seeds         = 1:20,
    bootstrap     = FALSE,
    parallel      = FALSE,
    show_progress = FALSE,
    config        = list(ci_method = "none")
  )
}

res_lm     <- run_fast("lm_att")
res_oracle <- run_fast("oracle_att")

summary_tbl <- dplyr::bind_rows(res_lm, res_oracle) %>%
  cs_tidy() %>%
  dplyr::group_by(estimator_id) %>%
  dplyr::summarise(
    n_runs = dplyr::n(),
    median_anchor_deviation = stats::median(att_error, na.rm = TRUE),
    median_abs_anchor_deviation = stats::median(abs(att_error), na.rm = TRUE),
    p90_abs_anchor_deviation = stats::quantile(abs(att_error), 0.90, names = FALSE, na.rm = TRUE),
    max_abs_anchor_deviation = max(abs(att_error), na.rm = TRUE),
    .groups   = "drop"
  )

knitr::kable(
  summary_tbl,
  digits = 3,
  caption = "Finite-replicate diagnostics against the structural signal anchor (n=1000, seeds=1:20)"
)
```

**Interpretation:** OLS exhibits non-convergence and extreme seed sensitivity
under the Cauchy mixture. Mean bias, RMSE, conventional coverage, and winner
rankings are not defined here. The structural oracle reproduces the governed
signal anchor; this confirms scorer consistency, not existence of a conventional
mean ATT.

## 8. Failure Mode Summary

- **L2/mean-based ATT estimators:** run them to diagnose non-convergence and
  extreme sensitivity; do not rank them or report conventional bias/RMSE/coverage.
- **Structural `true_att`:** a governed finite-sample signal anchor for
  diagnostics, not a conventional superpopulation mean potential-outcome ATT in
  this no-mean regime.
- **QST/quantile targets:** well-defined under the Cauchy mixture and the valid
  distributional comparison surface.

## 9. Implementation Reference

- Generator: `R/dgp-synth-heavytail.R`
- Oracle truth: `cs_get_oracle_qst("synth_heavytail", version = "1.6.0")`
- Metadata: `inst/dgp_meta/synth_heavytail.yml`

## 10. Validation Checklist

- [x] Mixture noise matches spec (0.8 * Normal + 0.2 * Cauchy).
- [x] Finite-sample structural signal anchor defined on `tau(X)` and
  distinguished from a conventional superpopulation mean potential-outcome ATT.
- [x] Visual diagnostics show extreme tails.
- [x] OLS non-convergence and extreme sensitivity reproduced.
- [x] Structural oracle reproduces the governed signal anchor.

## 11. Changelog

- v1.3.0 (2026-01-06): Initial dossier for heavy-tail stress DGP.
- v1.6.0: Implemented Common Random Numbers (CRN) for Oracle generation. This reduces Monte Carlo variance of the QST contrast; it does not eliminate empirical-quantile sampling uncertainty. Also enforces regime consistency (Normal/Cauchy) for counterfactuals during Oracle evaluation.

## Appendix: Implementation

```{r impl_code, echo=FALSE, comment=''}
# Dynamically retrieve the specific versioned function for this dossier
dgp_base <- paste0("dgp_", params$dgp_id)
dgp_ver  <- if (!is.null(params$meta$version)) {
  paste0(dgp_base, "_v", gsub("\\.", "", params$meta$version))
} else {
  dgp_base
}

if (exists(dgp_ver)) {
  cat(sprintf("Source code for: %s\n\n", dgp_ver))
  print(get(dgp_ver))
} else if (exists(dgp_base)) {
  cat(sprintf("Source code for: %s (Wrapper)\n\n", dgp_base))
  print(get(dgp_base))
} else {
  cat("Source code not found in loaded namespace.")
}
```

