---
title: "DGP Dossier: `r params$meta$dgp_id`"
output: html_document
params:
  meta: NULL
  dgp_id: "synth_hd_sparse_plm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(CausalStress)
library(ggplot2)
library(dplyr)
have_kableextra <- requireNamespace("kableExtra", quietly = TRUE)
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork_available <- TRUE
} else {
  patchwork_available <- FALSE
}
```

## 1. Identity & Status

```{r meta_table, echo=FALSE, results='asis'}
stars <- params$meta$difficulty$stars %||% 0
active_star   <- '<span style="color: #f1c40f;">&#9733;</span>'
inactive_star <- '<span style="color: #e0e0e0;">&#9734;</span>'
difficulty_val <- paste0(
  strrep(active_star, max(0, stars)),
  strrep(inactive_star, max(0, 5 - stars)),
  " (", stars, "/5)"
)
prof <- params$meta$stress_profile

get_badge_color <- function(val) {
  val <- tolower(as.character(val))
  if (val %in% c("linear", "gaussian", "strong", "high", "easy", "stable", "att")) return("success")
  if (val %in% c("moderate", "medium", "experimental")) return("warning")
  if (val %in% c("nonlinear", "heavytail", "heavy", "poor", "severe", "deprecated", "heterogeneous")) return("danger")
  "secondary"
}

stress_val <- if (is.null(prof)) {
  "N/A"
} else {
  badge_vec <- mapply(
    function(k, v) {
      cls <- get_badge_color(v)
      sprintf('<span class="badge bg-%s" style="margin-right:4px;">%s: %s</span>', cls, k, v)
    },
    names(prof),
    unlist(prof),
    USE.NAMES = FALSE
  )
  paste0("&nbsp;", paste(badge_vec, collapse = ""))
}

status_raw <- params$meta$status %||% "N/A"
status_class <- get_badge_color(status_raw)
status_val <- sprintf('<span class="badge bg-%s">%s</span>', status_class, status_raw)

version_val <- params$meta$version %||% "N/A"
dgp_id_val <- params$meta$dgp_id %||% "N/A"
if (is.character(dgp_id_val)) {
  dgp_id_val <- dgp_id_val
}

meta_df <- tibble::tibble(
  Attribute = c("DGP ID", "Version", "Status", "Difficulty", "Stress Profile"),
  Value = c(
    dgp_id_val,
    version_val,
    status_val,
    difficulty_val,
    stress_val
  )
)

knitr::kable(meta_df, format = "html", col.names = NULL, escape = FALSE) %>%
  {
    if (have_kableextra) {
      kableExtra::kable_styling(
        .,
        bootstrap_options = c("hover", "condensed"),
        full_width = TRUE,
        position = "left"
      ) %>%
        kableExtra::column_spec(1, bold = TRUE, width = "150px")
    } else {
      .
    }
  }
```

## 2. What This DGP Stresses (Intent)

Sparsity and regularization in high dimensions. Tests whether estimators can
recover a sparse linear signal without overfitting. Toeplitz correlation
(\(\rho = 0.95\)) increases collinearity and makes nuisance learning fragile,
but overlap is restored to be moderate in v1.5.0 so the primary stress is
high-dimensional sparsity rather than positivity.

## 3. Identification Assumptions (Explicit)

- Selection on observables holds.
- Overlap is moderate.
- SUTVA holds.

## 4. Mathematical Specification

Covariates:
\[
X \in \mathbb{R}^{100},\quad X \sim \mathcal{N}(0, \Sigma),\quad \Sigma_{ij} = 0.95^{|i-j|}
\]

Outcome (control):
\[
\mu_0(X) = \sum_{j=1}^{5} X_j
\]

Propensity:
\[
\text{logit}(p(X)) = 0.2\sum_{j=1}^{5} X_j
\]

Treatment effect:
\[
\tau(X) \equiv 1
\]

Noise:
\[
\varepsilon \sim \mathcal{N}(0, 1^2)
\]

## 5. Oracle Truth Definition

- True ATT is computed from `structural_te` among treated units.
- True QST is computed on the oracle tau grid (`cs_get_oracle_qst()`).

## 6. Visual Diagnostics (n = 5000)

```{r visuals, fig.width=10, fig.height=7}
set.seed(123)
dgp <- dgp_synth_hd_sparse_plm(n = 5000, seed = 123)
df <- dgp$df

p1 <- ggplot(df, aes(x = p, fill = factor(w))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(title = "Propensity Distribution (Moderate Overlap)",
       x = "p(X)", y = "Count", fill = "W") +
  theme_minimal()

p2 <- ggplot(df, aes(x = X1, y = structural_te)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Treatment Effect (Constant)",
       x = "X1", y = "structural_te") +
  theme_minimal()

covars <- paste0("X", 1:100)
std_mean_diff <- vapply(covars, function(x) {
  xval <- df[[x]]
  num <- abs(mean(xval[df$w == 1]) - mean(xval[df$w == 0]))
  den <- stats::sd(xval)
  if (is.na(den) || den == 0) return(0)
  num / den
}, numeric(1))
love_tbl <- data.frame(
  covariate = covars,
  smd = std_mean_diff,
  stringsAsFactors = FALSE
)
love_tbl <- love_tbl[order(love_tbl$smd, decreasing = TRUE), ]
love_top <- head(love_tbl, 10)

p3 <- ggplot(love_top, aes(x = smd, y = reorder(covariate, smd))) +
  geom_point(color = "#2c3e50", size = 2) +
  labs(
    title = "Love Plot (Top 10 Absolute SMDs)",
    x = "Absolute SMD",
    y = "Covariate"
  ) +
  theme_minimal()

p4 <- ggplot(dgp$true_qst, aes(x = tau, y = value)) +
  geom_hline(yintercept = dgp$true_att, linetype = "dashed", color = "gray70") +
  geom_line(color = "#2c3e50", size = 1) +
  labs(
    title = "True Quantile Shift (QST)",
    subtitle = "Constant effect under high-dimensional sparsity",
    x = "Quantile (tau)",
    y = "Treatment Effect"
  ) +
  theme_minimal()

if (patchwork_available) {
  library(patchwork)
  (p1 + p2) / (p3 + p4)
} else {
  print(p1)
  print(p2)
  print(p3)
  print(p4)
}
```

## 7. Empirical Validation

`lm_att` serves as the baseline: it matches the sparse linear outcome model.

**Victim:** `ipw_att` uses standard (unregularized) logistic regression for the propensity model. In $P=100$ with $\rho=0.95$, collinearity and low N/P can still make nuisance estimation fragile, but v1.5.0 restores moderate overlap so the primary stress is high-dimensional sparsity rather than positivity.

| estimator_id | mean_bias | rmse  |
|:-------------|:----------|:------|
| ipw_att      | -0.166    | 1.105 |
| lm_att       | 0.032     | 0.118 |
| oracle_att   | 0.000     | 0.000 |

## 8. Failure Mode Summary

- High-dimensional nuisance estimation can drown out the sparse signal.
- Non-parametric methods that fit complex nuisance surfaces can degrade sharply in 100D.

## 9. Implementation Reference

- Generator: `R/dgp-synth-hd-sparse-plm.R`
- Registry: `cs_dgp_registry()` entry for `synth_hd_sparse_plm`
- Oracle truth: `cs_get_oracle_qst("synth_hd_sparse_plm", version = "1.5.0")`

## 10. Validation Checklist

- [x] Toeplitz covariance specified
- [x] Sparse outcome/propensity specified
- [x] Love plot replaces X1 vs X2 scatter

## 11. Changelog

- v1.4.0: Parameter hardening (Deprecated v1.3.0).
- v1.5.0: Reduced propensity coefficients to 0.2. v1.4.0 deprecated due to unintended positivity violation caused by correlation inflation.

## Appendix: Implementation

```{r impl_code, echo=FALSE, comment=''}
# Dynamically retrieve the specific versioned function for this dossier
dgp_base <- paste0("dgp_", params$dgp_id)
dgp_ver  <- if (!is.null(params$meta$version)) {
  paste0(dgp_base, "_v", gsub("\\.", "", params$meta$version))
} else {
  dgp_base
}

if (exists(dgp_ver)) {
  cat(sprintf("Source code for: %s\n\n", dgp_ver))
  print(get(dgp_ver))
} else if (exists(dgp_base)) {
  cat(sprintf("Source code for: %s (Wrapper)\n\n", dgp_base))
  print(get(dgp_base))
} else {
  cat("Source code not found in loaded namespace.")
}
```
