---
title: "DGP Dossier: `r params$meta$dgp_id`"
output: html_document
params:
  meta: NULL
  dgp_id: "synth_placebo_tilted"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(CausalStress)
library(ggplot2)
have_kableextra <- requireNamespace("kableExtra", quietly = TRUE)
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork_available <- TRUE
} else {
  patchwork_available <- FALSE
}
```

## 1. Identity & Status

```{r meta_table, echo=FALSE, results='asis'}
stars <- params$meta$difficulty$stars %||% 0
active_star   <- '<span style="color: #f1c40f;">&#9733;</span>'
inactive_star <- '<span style="color: #e0e0e0;">&#9734;</span>'
difficulty_val <- paste0(
  strrep(active_star, max(0, stars)),
  strrep(inactive_star, max(0, 5 - stars)),
  " (", stars, "/5)"
)
prof <- params$meta$stress_profile

get_badge_color <- function(val) {
  val <- tolower(as.character(val))
  if (val %in% c("linear", "gaussian", "strong", "high", "easy", "stable", "att")) return("success")
  if (val %in% c("moderate", "medium", "experimental")) return("warning")
  if (val %in% c("nonlinear", "heavytail", "heavy", "poor", "severe", "deprecated", "heterogeneous")) return("danger")
  "secondary"
}

stress_val <- if (is.null(prof)) {
  "N/A"
} else {
  badge_vec <- mapply(
    function(k, v) {
      cls <- get_badge_color(v)
      sprintf('<span class="badge bg-%s" style="margin-right:4px;">%s: %s</span>', cls, k, v)
    },
    names(prof),
    unlist(prof),
    USE.NAMES = FALSE
  )
  paste0("&nbsp;", paste(badge_vec, collapse = ""))
}

status_raw <- params$meta$status %||% "N/A"
status_class <- get_badge_color(status_raw)
status_val <- sprintf('<span class="badge bg-%s">%s</span>', status_class, status_raw)

version_val <- params$meta$version %||% "N/A"
dgp_id_val <- params$meta$dgp_id %||% "N/A"
if (is.character(dgp_id_val)) {
  dgp_id_val <- dgp_id_val
}

meta_df <- tibble::tibble(
  Attribute = c("DGP ID", "Version", "Status", "Difficulty", "Stress Profile"),
  Value = c(
    dgp_id_val,
    version_val,
    status_val,
    difficulty_val,
    stress_val
  )
)

knitr::kable(meta_df, format = "html", col.names = NULL, escape = FALSE) %>%
  {
    if (have_kableextra) {
      kableExtra::kable_styling(
        .,
        bootstrap_options = c("hover", "condensed"),
        full_width = TRUE,
        position = "left"
      ) %>%
        kableExtra::column_spec(1, bold = TRUE, width = "150px")
    } else {
      .
    }
  }
```
## 2. What This DGP Stresses (Intent)

**Intent:** The "Spurious Correlation Trap."
This DGP induces strong (but manageable) linear confounding where treated units have naturally higher outcomes than control units ($p(X)$ aligns with $\mu_0(X)$).

**The Scientific Question:** "Can the estimator distinguish Selection Bias from Treatment Effect?"
A naive comparison suggests a strong positive effect. The estimator must vigorously adjust for covariates to recover the true zero. This tests resistance to "Bias Leakage."

## 3. Identification Assumptions (Explicit)

-   Selection on observables holds.
-   Overlap is moderately tilted (propensity shifted, but not extreme).
-   SUTVA holds.
-   Sharp Null: $Y_1 \equiv Y_0$.

## 4. Mathematical Specification

Covariates: $X_1, X_2, X_3, X_4, X_5 \sim \mathcal{N}(0, 1^2)$ independently.

**Outcome (control):** $$\mu_0(X) = 1 + X_1 + 0.5 X_2$$

**Propensity (tilted):** $$p(X) = \text{expit}(0.6 X_1 + 0.8 X_2)$$

**Treatment effect (sharp null):** $$\tau(X) \equiv 0$$

**Noise:** $$\varepsilon \sim \mathcal{N}(0, 0.5^2)$$ (Shared noise $Y_1 \equiv Y_0$)

## 5. Oracle Truth Definition

-   True ATT is zero by construction.
-   True QST is identically zero on the canonical grid.

## 6. Visual Diagnostics (n = 5000)

```{r visuals, fig.width=10, fig.height=7}
set.seed(123)
dgp <- dgp_synth_placebo_tilted(n = 5000, seed = 123)
df <- dgp$df

p1 <- ggplot(df, aes(x = p, fill = factor(w))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(title = "Propensity Distribution (Moderate Tilt)",
       x = "p(X)", y = "Count", fill = "W") +
  theme_minimal()

p2 <- ggplot(df, aes(x = X1, y = structural_te)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = expression(paste("Treatment Effect: ", tau(X), " = 0")),
       x = "X1", y = "structural_te") +
  theme_minimal()

p3 <- ggplot(df, aes(x = X1, y = X2, color = factor(w))) +
  geom_point(alpha = 0.4) +
  labs(title = "Covariate Overlap (Joint Support)", x = "X1", y = "X2", color = "W") +
  theme_minimal()

p4 <- ggplot(dgp$true_qst, aes(x = tau, y = value)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
  geom_line(color = "#2c3e50", size = 1) +
  labs(
    title = "True Quantile Shift (QST)",
    subtitle = "Sharp null: flat line at 0",
    x = "Quantile (tau)",
    y = "Treatment Effect"
  ) +
  theme_minimal()

if (patchwork_available) {
  library(patchwork)
  (p1 + p2) / (p3 + p4)
} else {
  print(p1)
  print(p2)
  print(p3)
  print(p4)
}
```

## 7. Empirical Validation

We validate only the oracle estimator to keep builds fast.

```{r empirical, cache=TRUE}
res_oracle <- cs_run_seeds(
  dgp_id        = "synth_placebo_tilted",
  estimator_id  = "oracle_att",
  n             = 1000,
  seeds         = 1:20,
  bootstrap     = FALSE,
  parallel      = FALSE,
  show_progress = FALSE
)

summary_tbl <- cs_tidy(res_oracle) %>%
  dplyr::summarise(
    mean_bias = mean(att_error, na.rm = TRUE),
    rmse      = sqrt(mean(att_error^2, na.rm = TRUE))
  )

knitr::kable(summary_tbl, digits = 4, caption = "Oracle validation (n=1000, seeds=1:20)")
```

## 8. Failure Mode Summary

-   **Bias Leakage:** The primary failure mode is reporting a positive effect (Type I error) because the estimator failed to fully scrub the confounding.
-   **Under-adjustment:** Regularized estimators (like Lasso) might over-shrink coefficients, leaving residual confounding that masquerades as a treatment effect.

## 9. Implementation Reference

-   Code: `R/dgp-synth-placebo-tilted.R` (v1.4.0)
-   Registry: `cs_dgp_registry()` entry for `synth_placebo_tilted`
-   Oracle: `true_att = 0`, `true_qst = 0`
-   Oracle QST: `cs_get_oracle_qst("synth_placebo_tilted", version = "1.4.0")`

## 10. Validation Checklist

-   [x] Sharp null enforced ($Y_1 \equiv Y_0$)
-   [x] Structural ATT = 0
-   [x] QST truth is identically zero
-   [x] Oracle validation passes

## 11. Changelog

-   v1.0 dossier: Initial placebo dossier
-   v1.4.0: Reduced tilt coefficients to 0.6/0.8. v1.3.0 deprecated to avoid confounding variance explosion with bias.

## Appendix: Implementation

```{r impl_code, echo=FALSE, comment=''}
# Dynamically retrieve the specific versioned function for this dossier
dgp_base <- paste0("dgp_", params$dgp_id)
dgp_ver  <- if (!is.null(params$meta$version)) {
  paste0(dgp_base, "_v", gsub("\\.", "", params$meta$version))
} else {
  dgp_base
}

if (exists(dgp_ver)) {
  cat(sprintf("Source code for: %s\n\n", dgp_ver))
  print(get(dgp_ver))
} else if (exists(dgp_base)) {
  cat(sprintf("Source code for: %s (Wrapper)\n\n", dgp_base))
  print(get(dgp_base))
} else {
  cat("Source code not found in loaded namespace.")
}
```



