---
title: "Benchmarking with suites and visualization"
format: html
vignette: >
  %\VignetteIndexEntry{Benchmarking with suites and visualization}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

Suites are named collections of DGP IDs. They make repeated benchmark setup
convenient, but membership is not evidence that a DGP is stable or that every
aggregate is scientifically meaningful. Inspect registry status and each DGP
report before designing a study.

```{r}
#| label: setup
#| message: false
library(CausalStress)
library(dplyr)
library(ggplot2)

board_path <- tempfile("causalstress-suite-")
dir.create(board_path)
board <- pins::board_folder(board_path)
```

## Discover suites

```{r}
cs_suite_registry() |>
  select(suite_id, description)

cs_get_suite("placebo")
```

The placebo suite currently contains experimental DGPs. The call below requests
that status explicitly and uses a small bootstrap only to make interval plots
executable. Increase the number of seeds and bootstrap replicates for scientific
work.

```{r}
#| warning: false
placebo_runs <- cs_run_suite(
  suite_id = "placebo",
  estimator_ids = "lm_att",
  n = 150,
  seeds = 1:2,
  status = "experimental",
  bootstrap = TRUE,
  B = 20,
  board = board,
  show_progress = FALSE
)

placebo_tidy <- cs_tidy(placebo_runs)
cs_collect_scores(placebo_runs) |>
  count(estimator_id, estimand_target_id, score_status)
```

## Visualize ATT evidence

`cs_plot_placebo()` shows each placebo ATT estimate and interval; color records
whether the interval covers the sharp-null target. `cs_plot_att_error()` shows
the finite set of per-run ATT deviations.

```{r}
#| fig-width: 7
#| fig-height: 4
cs_plot_placebo(placebo_tidy) +
  ggtitle("Placebo ATT intervals")
```

```{r}
#| fig-width: 7
#| fig-height: 4
cs_plot_att_error(placebo_tidy) +
  ggtitle("ATT deviations in the example runs")
```

## Interpret the gatekeeper components

`cs_summarise_gatekeeper()` returns separate target-specific components. The ATT
component compares interval coverage with `threshold`. ATE remains structurally
present but unverified pending recalibration. For QST, a run fails when more
than 10% of its tau intervals exclude the placebo truth, and an estimator fails
when more than 10% of verified runs fail. Missing intervals produce
`UNVERIFIED`, not a pass.

```{r}
gate <- cs_summarise_gatekeeper(placebo_runs, threshold = 0.90)

gate$att_verdict
gate$att_culprits
gate$ate_verdict
gate$qst_verdict
```

The two seeds above demonstrate the API; they are far too few to support a
scientific gatekeeper decision.

## Inspect execution metadata and persisted evidence

```{r}
placebo_tidy |>
  select(
    dgp_id, estimator_id, seed,
    run_time_dgp, run_time_est, run_time_total
  ) |>
  arrange(desc(run_time_total))

cs_audit(board) |>
  select(dgp_id, estimator_id, seed, fit_fingerprints)
```

Timing fields are provenance, not scientific identities. Persisted results are
written only because this example supplied `board`; `skip_existing = TRUE` is a
separate opt-in for strict fingerprint-checked resume.
