---
title: "Caching and strict resume"
format: html
vignette: >
  %\VignetteIndexEntry{Caching and strict resume}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

CausalStress can persist each completed seed to a pins board and later reuse
exactly matching artifacts. Persistence and resume are both opt-in. They reduce
the amount of completed work repeated after interruption; they cannot preserve
an estimator call that had not finished and been written.

```{r}
#| label: setup
#| message: false
library(CausalStress)
library(pins)
library(dplyr)

board_path <- tempfile("causalstress-resume-")
dir.create(board_path)
board <- board_folder(board_path)
```

## Persist completed seeds

Supplying `board` writes completed results as RDS-backed pins. Without a board,
the same runner returns results only.

```{r}
first_pass <- cs_run_seeds(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 100,
  seeds = 1:2,
  board = board,
  show_progress = FALSE
)

cs_audit(board) |>
  select(dgp_id, estimator_id, n, seed, fit_fingerprints)
```

## Resume with exact configuration identity

`skip_existing = TRUE` asks the runner to reuse completed schema-4 pins. For
each existing seed, it recomputes the expected configuration fingerprint and
requires an exact match. It then runs only seeds without a matching artifact
and returns all requested seeds in deterministic order.

```{r}
resumed <- cs_run_seeds(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 100,
  seeds = 1:3,
  board = board,
  skip_existing = TRUE,
  show_progress = FALSE
)

resumed |>
  select(seed, config_fingerprint)
```

Reading cached artifacts still has filesystem cost; “reused” does not mean
literally zero elapsed time. The important guarantee is that matching completed
seeds are not re-estimated.

## Configuration changes fail closed

Changing bootstrap settings, sample size, tau, estimator configuration,
versions, or another fingerprinted input means the old evidence is not the
requested evidence. Resume rejects the mismatch. Missing fingerprints and
historical fingerprint schemas are also rejected rather than silently migrated.

```{r}
mismatch <- tryCatch(
  cs_run_seeds(
    dgp_id = "synth_baseline",
    estimator_id = "lm_att",
    n = 100,
    seeds = 1:3,
    bootstrap = TRUE,
    B = 20,
    board = board,
    skip_existing = TRUE,
    show_progress = FALSE
  ),
  error = function(error) conditionMessage(error)
)

mismatch
```

This fail-closed check prevents an accidental mixed campaign under resume. It
does not claim that arbitrary artifacts outside the runner or deliberate manual
board edits can never be mixed.

## Recompute deliberately

Use `force = TRUE` or leave `skip_existing = FALSE` when replacement is
intentional. Existing matching pins are removed before the new result is
written.

```{r}
recomputed <- cs_run_seeds(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 100,
  seeds = 1:3,
  bootstrap = TRUE,
  B = 20,
  board = board,
  force = TRUE,
  show_progress = FALSE
)

recomputed |>
  select(seed, att_ci_width)
```

Deletion helpers are available when removing evidence is the actual intent:

```{r}
#| eval: false
cs_delete_result(
  board,
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 100,
  seed = 1
)

cs_delete_campaign(board, "synth_baseline", "lm_att")
```

## Parallel staging is a separate boundary

Experimental parallel workers do not write directly to a shared board. They
write RDS staging files, and the controlling process gathers them through a
single-writer path. This protects the board-write boundary and enables recovery
of completed staged results; it is not a transaction over a whole campaign.

| Intent | Setting |
|---|---|
| Return results without persistence | omit `board` |
| Persist newly completed results | `board = my_board` |
| Reuse exactly matching schema-4 results | `skip_existing = TRUE` |
| Recompute and replace requested results | `force = TRUE` |
| Inspect persisted records | `cs_audit(my_board)` |
