---
title: "Canonical workflow: from a governed run to typed scores"
format: html
vignette: >
  %\VignetteIndexEntry{Canonical workflow: from a governed run to typed scores}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

This guide is the ordinary CausalStress workflow. It uses exported functions
and estimators shipped with the package. A run produces estimator outputs and a
typed score surface; `cs_collect_scores()` is the canonical way to collect that
surface. The older ATT- and QST-specific collectors remain compatibility
projections.

```{r}
#| label: setup
#| message: false
library(CausalStress)
library(dplyr)
library(ggplot2)
```

## Discover governed components

Suites are named collections, not a scientific endorsement of every member.
Inspect the DGP status and estimator requirements before running a study.

```{r}
suites <- cs_suite_registry()
suites[, c("suite_id", "description")]

baseline <- cs_get_dgp("synth_baseline")
baseline[c("dgp_id", "version", "status", "type")]

lm_descriptor <- cs_get_estimator("lm_att")
lm_descriptor[c(
  "estimator_id", "type", "oracle", "supports_qst", "version",
  "source", "requires_pkgs"
)]
```

## Run once, then collect typed scores

`cs_run_single()` is the smallest ordinary entry point. The result keeps typed
estimator `outputs`, canonical `scores`, compatibility projections (`att` and
`qst`), bootstrap draws when requested, metadata, and provenance.

```{r}
one_run <- cs_run_single(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 200,
  seed = 17
)

names(one_run)

one_scores <- cs_collect_scores(one_run)
one_scores |>
  select(
    estimand_target_id, metric_id, score_status,
    estimate, truth, error, non_comparable_reason
  )
```

Each score row has an `estimand_target_id`. CausalStress compares an estimator
output only with truth for that same target:

- **ATT** is the mean structural effect among treated units when that mean is a
  valid target for the DGP regime.
- **ATE** is the finite-sample mean structural effect over all generated units.
- **QST** is a quantile-specific distributional contrast among treated units,
  indexed by `tau`.
- **CATE** is an explicitly staged target. Wave 1 records whether an estimator
  produced it, but does not yet score a CATE curve.

Targets are never silently cross-scored. A CATE estimate is not an ATT estimate,
and an ATT estimate is not a QST estimate.

## Request ATT and ATE explicitly

The built-in oracle is useful for testing benchmark truth plumbing. It is not
an ordinary estimator: its descriptor permits the runner to grant governed
`structural_te` through the Airlock.

```{r}
oracle_run <- cs_run_single(
  dgp_id = "synth_baseline",
  estimator_id = "oracle_att",
  n = 200,
  seed = 17,
  config = list(estimand_targets = c("att", "ate"))
)

cs_collect_scores(oracle_run) |>
  select(estimand_target_id, metric_id, score_status, estimate, truth, error)
```

QST truth is defined on potential-outcome distributions. This direct DGP call
is a truth-definition demonstration; do not pass `y0`, `y1`, `p`, or
`structural_te` to an ordinary estimator.

```{r}
truth_example <- dgp_synth_baseline(n = 300, seed = 19)
cs_true_qst(
  y0 = truth_example$df$y0,
  y1 = truth_example$df$y1,
  w = truth_example$df$w,
  tau = c(0.25, 0.50, 0.75)
)
```

No dependency-free estimator shipped with CausalStress currently produces
QST. The optional GenGC estimator does. Its example is deliberately opt-in so
the ordinary article remains executable without optional estimator packages.

```{r}
#| label: optional-qst-status
run_optional_qst <- requireNamespace("GenGC", quietly = TRUE) &&
  identical(tolower(Sys.getenv("CAUSALSTRESS_RUN_OPTIONAL_DOCS")), "true")

if (!run_optional_qst) {
  message(
    "Optional QST example skipped. Install GenGC and set ",
    "CAUSALSTRESS_RUN_OPTIONAL_DOCS=true to run it."
  )
}
```

```{r}
#| label: optional-qst-run
#| eval: !expr run_optional_qst
qst_run <- cs_run_single(
  dgp_id = "synth_qte1",
  estimator_id = "gengc",
  n = 300,
  seed = 19,
  status = "experimental",
  tau = c(0.25, 0.50, 0.75),
  config = list(estimand_targets = "qst")
)

cs_collect_scores(qst_run) |>
  select(estimand_target_id, tau, metric_id, score_status, estimate, truth, error)
```

## Make staged targets visible

A mixed request can preserve executable ATT evidence while explicitly showing
that CATE scoring is not implemented. The CATE row is evidence about the
current contract; it is not an ATT result under another name.

```{r}
staged <- cs_run_single(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 200,
  seed = 23,
  config = list(estimand_targets = c("att", "cate"))
)

cs_collect_scores(staged) |>
  select(
    estimand_target_id, metric_id, score_status,
    estimate, non_comparable_reason
  )
```

## Repeat over seeds and summarize

Use `cs_run_seeds()` for one DGP-estimator pair and `cs_run_grid()` for ordinary
cross-product work. Keep typed scores for target-aware analysis; use the
summary helpers only where the target and moment regime make the aggregate
meaningful.

```{r}
runs <- cs_run_grid(
  dgp_ids = "synth_baseline",
  estimator_ids = c("lm_att", "ipw_att"),
  n = 200,
  seeds = 1:3,
  show_progress = FALSE
)

typed_scores <- cs_collect_scores(runs)
typed_scores |>
  count(estimator_id, estimand_target_id, score_status)

summary <- cs_summarise_runs(runs)
summary
```

```{r}
#| fig-width: 7
#| fig-height: 4
cs_plot_att_error(runs)
```

Do not use mean-based ATT RMSE, coverage, or rankings for
`synth_heavytail`. That DGP intentionally probes an estimand boundary under
Cauchy-mixture noise: its governed structural contrast is a signal anchor, not
a conventional mean potential-outcome ATT. Run ATT estimators there to study
breakdown, but use QST for valid distributional comparisons rather than an ATT
shootout.

## Inspect provenance

Scientific identities live in the result metadata and score records;
non-scientific execution details live in provenance.

```{r}
cs_meta_flatten(one_run) |>
  select(
    dgp_id, dgp_version, estimator_id, estimator_version,
    config_fingerprint, fit_fingerprint
  )

cs_provenance(one_run)[c("run_time_dgp", "run_time_est", "run_time_total")]
```

## Persist completed runs and resume strictly

Persistence is opt-in. Without a `board`, results are returned but not stored.
With a pins board, completed results are written as RDS-backed pins. Resume is
also opt-in: `skip_existing = TRUE` reuses only schema-4 artifacts whose
configuration fingerprint matches exactly. A missing/old schema or a mismatch
fails closed. `force = TRUE` or `skip_existing = FALSE` recomputes instead.

```{r}
board_path <- tempfile("causalstress-board-")
dir.create(board_path)
board <- pins::board_folder(board_path)

first_pass <- cs_run_seeds(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 150,
  seeds = 1:2,
  board = board,
  show_progress = FALSE
)

resumed <- cs_run_seeds(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 150,
  seeds = 1:3,
  board = board,
  skip_existing = TRUE,
  show_progress = FALSE
)

resumed |>
  select(seed, config_fingerprint)

cs_audit(board) |>
  select(dgp_id, estimator_id, n, seed)
```

Persistence protects completed artifacts, not an in-progress estimator call or
an entire campaign transaction. Experimental parallel persistence additionally
uses worker staging followed by controlled single-writer gathering.

## Advanced planned batching

`cs_plan_campaign()` plus `cs_run_batch()` is the advanced lifecycle for
explicit plans and batch artifacts. It is distinct from ordinary single, seed,
and grid execution. `cs_run_campaign()` currently supports both a direct grid
mode and a planned-batch mode; this guide does not redefine or deprecate either
public contract.

```{r}
#| eval: false
plan <- cs_plan_campaign(
  dgp_list = c("synth_baseline"),
  estimator_list = c("lm_att", "ipw_att"),
  n_seeds = 1:100,
  batch_size = 20,
  strategy_map = list(defaults = list(n = 1000))
)

batch_1 <- cs_run_batch(
  plan = plan,
  batch_id = 1,
  staging_dir = "staging_batches"
)
```
