---
title: "Distributional intelligence with QST"
format: html
vignette: >
  %\VignetteIndexEntry{Distributional intelligence with QST}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

A mean effect and a distributional contrast answer different questions. The
experimental `synth_qte1` DGP has a sign-changing structural effect. Treatment
selection changes the composition of the treated group, so a mean ATT summary
cannot by itself describe the lower and upper parts of the treated
potential-outcome distributions.

```{r}
#| label: setup
#| message: false
library(CausalStress)
library(dplyr)
library(ggplot2)
```

## Inspect the governed truth

Direct DGP output contains potential outcomes for truth construction. Those
columns are runner-owned truth and are not inputs to an ordinary estimator.

```{r}
dgp <- dgp_synth_qte1(n = 1000, seed = 42)

tibble::tibble(
  structural_effect = dgp$meta$structural_te,
  treated = dgp$df$w
) |>
  count(treated, structural_effect)

cs_true_qst(
  y0 = dgp$df$y0,
  y1 = dgp$df$y1,
  w = dgp$df$w,
  tau = c(0.10, 0.25, 0.50, 0.75, 0.90)
)
```

## ATT-only evidence

The core linear estimator produces ATT only. Its typed score surface says that
explicitly; the scalar result must not be presented as a recovered QST curve.

```{r}
#| warning: false
att_run <- cs_run_single(
  dgp_id = "synth_qte1",
  estimator_id = "lm_att",
  n = 500,
  seed = 1,
  status = "experimental"
)

cs_collect_scores(att_run) |>
  select(estimand_target_id, metric_id, score_status, estimate, truth, error)
```

## Optional QST estimation

The shipped GenGC adapter is optional. The example runs only when GenGC is
installed and the environment variable `CAUSALSTRESS_RUN_OPTIONAL_DOCS=true`
is set. This guard is visible so a rendered article cannot make a skipped
optional analysis look like successful evidence.

```{r}
#| label: optional-gengc-status
run_optional_gengc <- requireNamespace("GenGC", quietly = TRUE) &&
  identical(tolower(Sys.getenv("CAUSALSTRESS_RUN_OPTIONAL_DOCS")), "true")

if (!run_optional_gengc) {
  message(
    "Optional GenGC example skipped. Install GenGC and set ",
    "CAUSALSTRESS_RUN_OPTIONAL_DOCS=true to execute it."
  )
}
```

```{r}
#| label: optional-gengc-run
#| eval: !expr run_optional_gengc
#| warning: false
qst_run <- cs_run_single(
  dgp_id = "synth_qte1",
  estimator_id = "gengc",
  n = 500,
  seed = 1,
  status = "experimental",
  tau = c(0.10, 0.25, 0.50, 0.75, 0.90),
  config = list(estimand_targets = "qst")
)

qst_scores <- cs_collect_scores(qst_run)
qst_scores |>
  select(estimand_target_id, tau, metric_id, score_status, estimate, truth, error)

cs_plot_qst(cs_collect_qst(cs_tidy(qst_run)))
```

QST estimates below zero at lower tau and above zero at higher tau would be
evidence of a sign-changing distributional contrast. The article does not claim
that result when the optional estimator run is skipped.

## Placebo gatekeeper

For QST placebo evidence, a run fails when more than 10% of verified tau
intervals exclude the null. An estimator fails when more than 10% of its
verified runs fail. ATT uses its separately supplied coverage threshold; the ATE
gatekeeper component is currently deferred and returns `UNVERIFIED`.

```{r}
#| eval: false
placebo_runs <- cs_run_suite(
  suite_id = "placebo",
  estimator_ids = "gengc",
  n = 1000,
  seeds = 1:100,
  status = "experimental",
  bootstrap = TRUE,
  B = 500
)

cs_summarise_gatekeeper(placebo_runs)
```

The gatekeeper needs interval-bearing placebo runs and enough replicates for its
decision; a point-estimate example is not a pass.
