---
title: "Extend CausalStress with an estimator"
format: html
vignette: >
  %\VignetteIndexEntry{Extend CausalStress with an estimator}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

`cs_register_estimator()` is the supported runtime extension boundary for
estimators. Registration is process-local: repeat it in each R session, or have
an extension package register during its own startup. CausalStress does not
persist registrations, replace an existing ID, or provide an unregister
operation.

```{r}
#| label: setup
#| message: false
library(CausalStress)
library(dplyr)
```

## A minimal typed estimator

An estimator generator accepts `(df, config = list(), tau = cs_tau_oracle,
...)`. The `df` supplied to an ordinary estimator contains observed `y`, binary
`w`, and covariates. This deliberately simple unadjusted mean difference is a
contract example, not a generally adequate causal estimator.

```{r}
mean_difference_att <- function(
  df,
  config = list(),
  tau = cs_tau_oracle,
  ...
) {
  estimate <- mean(df$y[df$w == 1]) - mean(df$y[df$w == 0])

  list(
    outputs = list(
      att = list(estimate = estimate)
    ),
    meta = list(
      estimator_id = "mean_difference_att",
      version = "1.0.0",
      capabilities = "att"
    )
  )
}

cs_register_estimator(
  estimator_id = "mean_difference_att",
  type = "external",
  generator = mean_difference_att,
  version = "1.0.0",
  description = "Unadjusted mean difference for contract documentation."
)

cs_get_estimator("mean_difference_att")[c(
  "estimator_id", "type", "version", "source", "requires_pkgs"
)]
```

The ordinary runner now treats the registered implementation like any other
estimator and scores only the requested target.

```{r}
custom_run <- cs_run_single(
  dgp_id = "synth_baseline",
  estimator_id = "mean_difference_att",
  n = 200,
  seed = 4
)

cs_collect_scores(custom_run) |>
  select(estimand_target_id, metric_id, score_status, estimate, truth, error)
```

## Treat identity and versioning as evidence

The estimator ID and version enter provenance and run identities. Change the
version whenever the scientific implementation changes. Because IDs cannot be
replaced within a session, restart R before registering a new implementation
under the same ID. For reproducible campaigns, define registration in a
versioned package or script and record the versions of its dependencies.

## Typed and legacy return shapes

Typed output is preferred. Each produced target is named under `outputs`:

```{r}
#| eval: false
list(
  outputs = list(
    att = list(estimate = 0.5),
    ate = list(estimate = 0.4),
    qst = tibble::tibble(
      tau = c(0.25, 0.50, 0.75),
      estimate = c(0.2, 0.4, 0.8)
    )
  ),
  meta = list(
    estimator_id = "my_estimator",
    version = "1.2.0",
    capabilities = c("att", "ate", "qst")
  )
)
```

QST output must contain the exact requested `tau` coordinates and an
`estimate` column. A legacy `list(att, qst, meta)` result is still normalized
for compatibility, but new extensions should return typed `outputs` so target
identity is explicit.

```{r}
#| eval: false
list(
  att = list(estimate = 0.5),
  qst = tibble::tibble(tau = tau, estimate = qst_estimates),
  meta = list(estimator_id = "legacy_shape", version = "1.0.0")
)
```

Registration's `supports_qst = TRUE` declares that the estimator can produce a
QST curve. Actual target production is still determined by the returned output
and the requested `config$estimand_targets`. It does not authorize substituting
QST for ATT, ATE, or CATE.

## Optional dependencies

Declare packages used by the implementation in `requires_pkgs`. The runner
checks them before calling the generator. If one is unavailable, the run
records a controlled estimator failure rather than entering the generator.

```{r}
#| eval: false
cs_register_estimator(
  estimator_id = "my_optional_estimator",
  type = "external",
  generator = my_optional_estimator,
  version = "1.0.0",
  requires_pkgs = c("someModelPackage")
)
```

Keep the package namespace explicit inside the generator, and test both the
installed and missing-dependency paths in the extension's CI.

## Confidence intervals

`bootstrap = TRUE` is a runner convenience: when absent, it supplies
`config$ci_method = "bootstrap"`, the per-run seed, and `config$n_boot` from
`B`. The estimator still owns interval computation and must return interval
fields and truthful CI metadata. Do not label a normal approximation, model
interval, or failed bootstrap as a successful bootstrap interval.

```{r}
#| eval: false
list(
  outputs = list(
    att = list(
      estimate = estimate,
      ci_lo = ci[1],
      ci_hi = ci[2]
    )
  ),
  meta = list(
    estimator_id = "my_estimator",
    version = "1.2.0",
    capabilities = "att",
    ci_method = "bootstrap",
    ci_type = "percentile",
    ci_level = 0.95,
    ci_valid = TRUE
  )
)
```

## The Airlock boundary

Ordinary estimators receive observed `y`, `w`, and covariates only. The runner
removes DGP-owned `y0`, `y1`, `p`, and `structural_te`. Only a registry
descriptor marked `oracle = TRUE` may be eligible for `p` or `structural_te`,
and the grant must also follow its declared columns and run configuration.
`y0` and `y1` are never granted. This is an honest-code scientific boundary,
not a security sandbox for hostile R code.

External estimators should normally remain non-oracle. Do not request oracle
metadata merely to simplify an implementation.

## Failures are classed at the boundary

Invalid registration and unknown or duplicate IDs use
`causalstress_registry_error`. Invalid estimator output uses estimator or
contract error classes. Truth-access violations use
`causalstress_airlock_error`. Consumers should catch the narrowest relevant
class rather than parse message text.

```{r}
duplicate_class <- tryCatch(
  {
    cs_register_estimator(
      estimator_id = "mean_difference_att",
      type = "external",
      generator = mean_difference_att
    )
    NA_character_
  },
  causalstress_registry_error = function(error) class(error)[1]
)

duplicate_class
```

## Extension-package checklist

Before using an external estimator in an evidence campaign:

1. freeze a unique ID and implementation version;
2. return explicit typed targets and exact requested QST coordinates;
3. declare and test optional dependencies;
4. test point estimates, interval metadata, deterministic seed forwarding, and
   classed failure paths;
5. verify ordinary execution cannot see DGP truth;
6. register afresh in every worker/session that needs the estimator; and
7. retain the registration code and dependency versions with the campaign.
