---
title: "from-run-to-history"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{from-run-to-history}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the **Constitutional Workflow** of CausalStress:

1.  **Run:** Execute simulations with strict guarantees.

2.  **Persist:** Save every seed atomically to disk.

3.  **Audit:** Query the history of your experiments.

4.  **Tidy:** Flatten complex results for analysis.

We will compare two estimators (`lm_att`, `ipw_att`) on two data-generating processes (`synth_baseline`, `synth_heavytail`).

## 1. Setup: The Environment

First, we load the package and set up a **Board**. In `CausalStress`, results are never just "returned" to the console—they are persisted to a storage backend (local folder, S3, etc.).

For this tutorial, we use a temporary board.

```{r setup}
library(CausalStress)
library(dplyr)
library(pins)

# In a real project, use board_folder("experiments") or board_s3(...)
board <- pins::board_temp()
```

## 2. The Atomic Building Block: `cs_run_single()`

The smallest unit of work in `CausalStress` is a single seed. This function returns a **Rich Object** containing the estimate, bootstrap distribution, and metadata.

```{r}
# Run one specific configuration
run_1 <- cs_run_single(
  dgp_id       = "synth_baseline",
  estimator_id = "lm_att",
  n            = 500,
  seed         = 123,
  bootstrap    = TRUE,
  B            = 100,
  board        = board  # <--- Persist this run immediately
)

# Inspect the Rich Object (It's a list)
str(run_1, max.level = 1)
```

While rich objects are great for storage, they are annoying for analysis. We use `cs_tidy()` to flatten them.

```{r}

run_1 %>%
cs_tidy() %>%
  select(dgp_id, estimator_id, seed, est_att, att_ci_width)

```

## 3. Scaling Up: `cs_run_grid()`

Real benchmarks involve running many seeds across many scenarios. `cs_run_grid()` orchestrates this.

Crucially, it supports **Atomic Persistence** and **Crash Recovery**.

```{r}

  # Define the benchmark grid
# 2 DGPs x 2 Estimators x 5 Seeds = 20 Runs
runs <- cs_run_grid(
  dgp_ids       = c("synth_baseline", "synth_heavytail"),
  estimator_ids = c("lm_att", "ipw_att"),
  n             = 500,
  seeds         = 1:5,
  bootstrap     = TRUE,
  B             = 100,
  board         = board,        # Save every seed
  skip_existing = TRUE          # Enable Crash Recovery
)
```

### The "Free Lunch" (Crash & Resume)

Because we set `skip_existing = TRUE`, we can re-run the exact same command, and `CausalStress` will detect that the work is already done.

```{r}


system.time({
  runs_resume <- cs_run_grid(
    dgp_ids       = c("synth_baseline", "synth_heavytail"),
    estimator_ids = c("lm_att", "ipw_att"),
    n             = 500,
    seeds         = 1:5,
    board         = board,
    skip_existing = TRUE
  )
})
```

The second run finishes almost instantly. In a real campaign on a cluster, this means you can interrupt and resume your work at any time without losing progress.

## 4. The Scorecard

Once the grid is done, we tidy the results and compute the benchmark metrics.

```{r}

  # 1. Tidy (Flatten lists to rows)
runs_tidy <- runs %>%
        cs_tidy()

# 2. Summarise (Compute Bias, RMSE, Coverage)
scorecard <- runs_tidy %>%
        cs_summarise_runs

scorecard %>%
        select(dgp_id, estimator_id, RMSE = mean_error, Coverage = mean_att_covered)
```

## 5. Provenance: The "Time Machine"

Imagine you ran an experiment last week, but you forgot which Git commit you used. Or you want to load a specific result from a month ago or a paper you published years ago.

`cs_audit()` turns your `pins` board into a searchable database of your research history.

```{r}
history <- cs_audit(board)

# What have we run so far?
history %>%
  select(dgp_id, estimator_id, seed, timestamp, git_hash) %>%
  head(5)
```

### Loading a specific past run

You can filter this history to find a specific artifact and load it back into R.

```{r}
# "Find that specific run I did for the heavy tail scenario..."
target <- history %>%
  filter(dgp_id == "synth_heavytail", estimator_id == "lm_att", seed == 1)

# Load the full rich object
old_result <- pins::pin_read(board, target$pin_name)

# Verify it matches
cs_tidy(old_result)$est_att
```

## Summary

This workflow ensures that your research is:

1.  **Safe:** No data leakage (Airlock).

2.  **Durable:** No lost work (Atomic Persistence).

3.  **Reproducible:** Exact provenance tracking (Audit).