---
title: "From a run to persisted history"
format: html
vignette: >
  %\VignetteIndexEntry{From a run to persisted history}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

CausalStress separates execution, optional persistence, typed scoring, and
audit. A board is required only when results should be stored; without it,
runners simply return their result objects.

```{r}
#| label: setup
#| message: false
library(CausalStress)
library(dplyr)
library(pins)

board_path <- tempfile("causalstress-history-")
dir.create(board_path)
board <- board_folder(board_path)
```

## Run and persist one seed

`cs_run_single()` returns typed outputs and scores, compatibility ATT/QST
projections, optional bootstrap draws, metadata, and provenance. Bootstrap draws
are present only when the estimator/runner produced them.

```{r}
one_run <- cs_run_single(
  dgp_id = "synth_baseline",
  estimator_id = "lm_att",
  n = 200,
  seed = 123,
  board = board
)

names(one_run)

cs_collect_scores(one_run) |>
  select(estimand_target_id, metric_id, score_status, estimate, truth, error)

cs_tidy(one_run) |>
  select(dgp_id, estimator_id, seed, est_att, att_error)
```

Only the completed result is persisted. The board does not make an in-progress
estimator call recoverable.

## Scale to an ordinary grid

This small grid uses the finite-moment baseline DGP, so the displayed mean ATT
error summaries have their ordinary interpretation. Heavy-tail estimand-boundary
runs should be executed and reported separately, not placed in a mean-based ATT
RMSE, coverage, or ranking shootout.

```{r}
runs <- cs_run_grid(
  dgp_ids = "synth_baseline",
  estimator_ids = c("lm_att", "ipw_att"),
  n = 200,
  seeds = 1:3,
  board = board,
  show_progress = FALSE
)

scorecard <- cs_summarise_runs(runs)
scorecard |>
  select(
    dgp_id, estimator_id, n_runs,
    mean_error, mean_abs_error, sd_error
  )
```

`mean_error` is signed mean deviation, `mean_abs_error` is mean absolute
deviation, and `sd_error` is the standard deviation across the supplied runs.
None of those columns is RMSE.

## Resume matching completed work

Resume is explicit and strict. `skip_existing = TRUE` requires current schema-4
metadata and an exact configuration fingerprint before reading an existing pin.
Here all requested records already exist, so the estimators are not rerun.

```{r}
resumed <- cs_run_grid(
  dgp_ids = "synth_baseline",
  estimator_ids = c("lm_att", "ipw_att"),
  n = 200,
  seeds = 1:3,
  board = board,
  skip_existing = TRUE,
  show_progress = FALSE
)

resumed |>
  count(estimator_id)
```

Filesystem reads and validation still take time. A mismatch or historical
schema fails closed; use `force = TRUE` only when recomputation and replacement
are intentional.

## Audit and reload persisted evidence

`cs_audit()` inventories pins and their recorded identities. It does not turn
timestamps or runtime logs into scientific identity fields.

```{r}
history <- cs_audit(board)
history |>
  select(
    dgp_id, estimator_id, estimator_version,
    seed, fit_fingerprints, score_fingerprints, pin_name
  ) |>
  head()
```

A pins board remains an ordinary persistence boundary, so a selected artifact
can be loaded with pins and then inspected through CausalStress accessors.

```{r}
target <- history |>
  filter(
    dgp_id == "synth_baseline",
    estimator_id == "lm_att",
    seed == 1
  ) |>
  slice(1)

stored_result <- pin_read(board, target$pin_name)

cs_meta_flatten(stored_result) |>
  select(dgp_id, estimator_id, seed, fit_fingerprint)

cs_provenance(stored_result)[c("run_time_dgp", "run_time_est", "run_time_total")]
```

For experimental parallel execution, workers stage completed RDS artifacts and
the controlling process gathers them through a single-writer path. That is a
different contract from serial board writes and does not make a whole campaign
transactional.
